# Metabolism
```{r, include=FALSE}
current_chapter <- "04-Metabolism.Rmd"
```
```{r, child="_download_button.Rmd"}
```

## Overview

Metabolism determines the baseline and dynamic energy demand required for fish to maintain physiological function, support movement, regulate ions, digest food, and sustain long-term migration. Within P-MEM, metabolic rate is computed as a function of body size, temperature, and age, following established allometric scaling relationships in ectothermic organisms, particularly fishes (e.g., [@kleiber_body_1932; @west_general_1997; @brown_toward_2004]). Temperature sensitivity is represented using a Q10 formulation consistent with ectothermic metabolic responses [@schmidt_nielsen_animal_1997], while age-dependent scaling reflects ontogenetic changes in metabolic intensity [@glazier_beyond_2005; @glazier_relevance_2023]. Metabolic efficiency, based on divergence from an agent’s optimal temperature, is then used to scale energy allocation across key internal processes including swimming, osmoregulation, and lipid catabolism.

## Purpose

To simulate energetically realistic metabolic costs for migratory fishes
in coastal and estuarine environments by accounting for size scaling,
temperature dependence, age effects, and dynamic energy partitioning
across physiological processes.

## Entities, State Variables, and Scales

### Spatial and Temporal Scales

-   **Spatial Unit**: Patch (3 m x 3 m resolution)
-   **Temporal Unit**: 5-minute time steps (`tick`)

### Global Variables

| Global Variable | Definition |
|------------------------------------------|------------------------------|
| **Q10** | Temperature coefficient used for metabolic scaling. |
| **base-constant** | Global constant for baseline metabolic scaling (0.1). |
| **age-sensitivity** | Scalar controlling how metabolism changes with age (0.5). |

### Patch Variables

| Variable Name | Definition |
|----------------------------------------|--------------------------------|
| **temperature** | The water temperature of the current patch used to calculate temperature dependent metabolic rate. |

### Agent Variables

| Variable Name | Definition |
|----------------------------------------|--------------------------------|
| **size** | Body mass or size index used for allometric scaling. |
| **age-num** | Age used in scaling metabolic rate. |
| **optimal-temperature** | Preferred temperature at which baseline metabolism is defined. |
| **max-survival-temp** | Maximum temperature at which the agent can survive and function. |
| **Met-base** | Baseline metabolic constant calculated during initialization. |
| **metabolism-rate** | Metabolic rate for the current tick. |
| **total-metabolism** | Accumulated metabolism over time. |
| **swim-efficiency** | Efficiency applied to movement when metabolic rate is lower or higher than baseline. |
| **osmoregulation-efficiency** | Efficiency applied to osmoregultion when metabolic rate is lower or higher than baseline. |
| **lipid-catabolism-efficiency** | Efficiency applied to lipid catabolism when metabolic rate is lower or higher than baseline. |
| **swim-base** | Baseline allocation of metabolism toward movement. |
| **E-base** | Base metabolic maintenance cost. |
| **E-creation** | Energy cost for creating new chloride cells. |
| **E-osmo** | Total energy spent on osmoregulation. |
| **motion-ratio** | Fraction of metabolism allocated to swimming. |
| **osmo-ratio** | Fraction of metabolism allocated to osmoregulation. |
| **M-max** | Maximum size for scaling across the population. |
| **max-age** | Maximum age for scaling across the population. |

## Process Overview and Scheduling

1.  Retrieve patch temperature (°C) and verify safe environmental
    inputs.

2.  Compute allometric, temperature, and age scalars (dimensionless).

3.  Calculate metabolic rate for the current tick ((energy units)
    tick⁻¹).

4.  Update metabolic accumulation (energy units).

5.  Compute a reference baseline metabolic rate ((energy units) tick⁻¹).

6.  Apply efficiency multipliers to swimming, digestion, osmoregulation,
    and lipid catabolism (%).

## Design Concepts

**Basic Principles:** Metabolic rate combines allometric scaling,
temperature responses, and age scaling consistent with metabolic ecology
theory (Brown et al., 2004; West et al., 1997; Kleiber, 1932). The model
uses a Q10 temperature function [@schmidt_nielsen_animal_1997],
Kleiber-type allometry [@kleiber_body_1932; @west_general_1997], and
age-dependent reduction in metabolic intensity [@glazier_beyond_2005;
@glazier_relevance_2023].

**Emergence:** Energetic constraints emerge from interactions among
size, temperature, and age. These constraints then shape movement,
digestion, and osmoregulation.

**Adaptation:** Agents adapt by adjusting internal efficiencies that
influence downstream behavioral and physiological submodels.

**Objectives:** Agents maintain adequate energy for survival and
migration and balance energetic cost across processes.

**Sensing:** Agents sense the temperature of the patch they occupy.

**Stochasticity:** The function itself is deterministic but stochastic
variation in individual size and age creates emergent metabolic
variation.

**Collectives:** Scaling uses the maximum size and age of the entire
population to normalize differences among individuals.

**Observation:** Outputs include metabolism-rate, total-metabolism, and
efficiency multipliers.

## Initialization

| Variable | Initialized Value | Justification |
|:----------------:|:----------------------------:|:----------------------:|
| Met-base | Computed using size, age, and temperature factors | Establishes a physiologically realistic baseline. |
| swim-base | Met-base times movement ratio | Allocates baseline metabolism for locomotion. |
| E-base | Met-base times 0.02 | Maintenance cost of cellular metabolic activity. |
| E-creation | Met-base times 0.05 | Cost of forming new chloride cells. |
| motion-ratio | 0.7 normalized | Movement is energetically dominant. |
| osmo-ratio | 0.3 normalized | Osmoregulation is energetically significant. |
| E-osmo | 0 | No osmoregulation cost at initialization. |
| metabolism-rate | Met-base | Baseline metabolism in the absence of environmental variation. |
| total-metabolism | 0 | Starts accumulation at zero. |

## Submodels

### Initialize Metabolism

Baseline metabolic constant is computed at initialization for each agent
using species-specific traits and fixed parameter values and remains
constant across simulation runs:

Baseline metabolic constant is computed using:

$$
Met_{base} = C_{base} \left(\frac{A_{mass}}{M_{max}}\right)^{0.75} \left[1 + S_{age}\left(1 - \frac{A_{age}}{Age_{max}}\right)\right] \left(\frac{T_{opt}}{T_{max}}\right) \hspace{2cm} \text{(C.1)}
$$

Where:

-   $Met_{base}$ is the baseline metabolic rate used as the reference
    for all energetic scaling ((energy units) tick⁻¹).

-   $C_{base}$ is the constant baseline coefficient for metabolic
    initialization, set as 0.1 (dimensionless).

-   $A_{mass}$ is an agent's body mass used for allometric and energetic
    scaling (g).

-   $M_{max}$ is the maximum body mass across the population (g).

-   $S_{age}$ is the sensitivity parameter controlling age-related
    metabolic scaling, set to 0.5 (dimensionless).

-   $A_{age}$ is the agent's age in years (years).

-   $Age_{max}$ is the maximum age in the population (years).

-   $T_{opt}$ is the optimal temperature for metabolic performance (°C).

-   $T_{max}$ is the maximum temperature for agent survival, used for
    normalization (°C).

### Energy Partitions

Baseline energy partitions are computed from the initialized metabolic
base:

$$
swim_{base} = Met_{base} \cdot Mt_{ratio}  \hspace{2cm} \text{(C.2)}
$$

Where:

-   $swim_{base}$ is the baseline energetic cost of swimming ((energy
    units) tick⁻¹).

-   $Met_{base}$ is the baseline metabolic rate used as the reference
    for all energetic scaling ((energy units) tick⁻¹).

-   $Mt_{ratio}$ is the ratio partitioning metabolic rate to baseline
    swimming cost (%).

$$
E_{\text{base}} = Met_{base} \cdot 0.02  \hspace{2cm} \text{(C.3)}
$$

Where:

-   $E_{base}$ is the baseline cell maintenance energy cost per tick
    ((energy units) tick⁻¹).

-   $Met_{base}$ is the baseline metabolic rate used as the reference
    for all energetic scaling ((energy units) tick⁻¹).

$$
E_{\text{creation}} = Met_{base} \cdot 0.05  \hspace{2cm} \text{(C.4)}
$$

Where:

-   $E_{creation}$ is the energy cost of chloride cell creation scaled
    by body size ((energy units) tick⁻¹).

-   $Met_{base}$ is the baseline metabolic rate used as the reference
    for all energetic scaling ((energy units) tick⁻¹).
    
### Metabolic Rate Calculation

At each tick, metabolism is recalculated using allometric, temperature,
and age scaling. The 0.75 exponent follows quarter-power scaling
relationships observed across taxa [@kleiber_body_1932;
@west_general_1997; @brown_toward_2004]. This formulation reflects
metabolic scaling relationships in teleost fishes and is applied
consistently across species, with species-specific differences
represented through parameter values (e.g., age, body mass, age
structure, and thermal preferences).

Size Factor:

$$
Size_{f} = \left(\frac{{A_{mass}}}{M_{max}}\right)^{0.75}  \hspace{2cm} \text{(C.5)}
$$

Where:

-   $Size_{f}$ is the allometric scaling factor derived from relative
    body mass (dimensionless).

-   $A_{mass}$ is the agent's body mass used for allometric and
    energetic scaling (g).

-   $M_{max}$ is the maximum body mass across the population (g).

The Q10 formulation is widely used to represent the temperature
sensitivity of metabolic rate in ectotherms
[@schmidt_nielsen_animal_1997; @brown_toward_2004].

The Temperature Factor is implemented as:

$$
T_{f} = Q_{10}^{\left(\frac{T_{\text{patch}} - T_{\text{opt}}}{10}\right)}  \hspace{2cm} \text{(C.6)}
$$

Where:

-   $T_{f}$ is the temperature scaling factor applied to metabolism
    (dimensionless).

-   $Q_{10}$ is the temperature sensitivity coefficient in the Q10
    formulation (dimensionless).

-   $T_{patch}$ is the local patch temperature experienced by the agent
    (°C).

-   $T_{opt}$ is the agent's optimal temperature for metabolic
    performance (°C).

Declines in mass-specific metabolic intensity with ontogeny and aging
are incorporated using the following age-scaling factor
[@glazier_beyond_2005; @glazier_relevance_2023]:

The Age Factor is calculated as:

$$
Age_{f} = 1 + \left(1 - \frac{A_{age}}{Age_{max}}\right)\cdot 0.5  \hspace{2cm} \text{(C.7)}
$$

Where:

-   \$\$Age\_{f}\$ is the age-based scaling factor modifying metabolic
    intensity (dimensionless).

-   $A_{age}$ is the agent's age in years (years).

-   $Age_{max}$ is the maximum age in the population (years).

This multiplicative formulation follows the structure of metabolic
ecology models in which metabolic rate scales as a function of body mass
and temperature [@brown_toward_2004].

Total metabolic rate:

$$
Met_{tick} = Met_{base} \cdot 
Size_{f} \cdot 
T_{f} \cdot 
Age_{f}  \hspace{2cm} \text{(C.8)}
$$

Where:

-   $Met_{tick}$ is the total metabolic rate for the current simulation
    tick ((energy units) tick⁻¹).

-   $Met_{base}$ is the baseline metabolic rate used as the reference
    for all energetic scaling ((energy units) tick⁻¹).

-   $Size_{f}$ is the allometric scaling factor derived from relative
    body mass (dimensionless).

-   $T_{f}$ is the temperature scaling factor applied to metabolism
    (dimensionless).

-   $Age_{f}$ is the age-based scaling factor modifying metabolic
    intensity (dimensionless).

Metabolic accumulation:

$$
Met_{total} = Met_{total} + Met_{tick}  \hspace{2cm} \text{(C.9)}
$$

Where:

-   $Met_{total}$ is the cumulative metabolic expenditure over time
    (energy units).

-   $Met_{tick}$ is the total metabolic rate for the current simulation
    tick ((energy units) tick⁻¹).
    
### Efficiency Multipliers

A baseline metabolic reference is defined using:

-   baseline size-factor = 1\
-   baseline age-factor = 1.5\
-   baseline temp-factor = 1

Thus:

$$
M_{\text{baseline}} =
\text{Met-base} \cdot 1 \cdot 1 \cdot 1.5
$$

Efficiency is computed as:

$$
Eff = 
\frac{Met_{tick}}{Met_{base}}  \hspace{2cm} \text{(C.10)}
$$

Where:

-   $Eff$ is the ratio of current to baseline metabolic rate
    (dimensionless).

-   $Met_{tick}$ is the total metabolic rate for the current simulation
    tick ((energy units) tick⁻¹).

-   $Met_{base}$ is the baseline metabolic rate used as the reference
    for all energetic scaling ((energy units) tick⁻¹).

This value is applied to:

-   $swim_{Eff}$ is the efficiency multiplier applied to swimming
    energetics (dimensionless).

-   $osmo_{Eff}$ is the efficiency multiplier applied to osmoregulatory
    costs (dimensionless).

-   $lipid_{Eff}$ is the efficiency multiplier applied to lipid
    catabolism (%).

Efficiency dynamically scales the cost of movement, ion regulation, and
lipid catabolism.

## Netlogo Implementation

```{netlogo}

;; ============================
;;   METABOLISM SUBMODEL
;; ============================

globals [
  ;; global physiological constants
  Q10
]

patches-own [
  temperature
]

fish-own [
  ;; core state variables
  weight
  age-num
  optimal-temperature
  max-survival-temp

  ;; metabolic scalars
  Met-base
  metabolism-rate
  total-metabolism

  ;; efficiency multipliers
  swim-efficiency
  osmoregulation-efficiency
  lipid-catabolism-efficiency

  ;; energy partitioning
  swim-base
  E-base
  E-creation
  E-osmo
  motion-ratio
  osmo-ratio

  ;; population normalization
  M-max
  max-age
]

;; ======================================
;;       INITIALIZE METABOLISM
;; ======================================

to initialize-metabolism
  let beta 0.75
  let age-sensitivity 0.5
  let base-constant 0.1
  
  ;; normalize population scaling
  set M-max max [weight] of fish
  set max-age max [age-num] of fish
  if M-max <= 0 [ set M-max 1 ]
  if max-age <= 0 [ set max-age 1 ]

  ;; scaling factors
  let size-factor (weight / M-max) ^ beta
  let age-factor 1 + age-sensitivity * (1 - (age-num / max-age))
  let temp-factor (optimal-temperature / max-survival-temp)

  ;; baseline metabolic constant
  set Met-base base-constant * size-factor * age-factor * temp-factor

  ;; energy partitions
  let motion-ratio-local 0.7
  let osmo-ratio-local   0.3
  let total-ratio (motion-ratio-local + osmo-ratio-local)

  set motion-ratio (motion-ratio-local / total-ratio)
  set osmo-ratio   (osmo-ratio-local / total-ratio)

  set swim-base (Met-base * motion-ratio)
  set E-osmo 0

  ;; fixed fractions
  set E-base     (Met-base * 0.02)
  set E-creation (Met-base * 0.05)

  ;; initial metabolic state
  set metabolism-rate Met-base
  set total-metabolism 0
end


;; ======================================
;;        CALCULATE METABOLISM EACH TICK
;; ======================================

to calculate-metabolism
  let beta 0.75

  ;; repopulate scaling values
  set M-max max [weight] of fish
  set max-age max [age-num] of fish
  if M-max <= 0 [ set M-max 1 ]
  if max-age <= 0 [ set max-age 1 ]

  ;; get patch temperature safely
  let T-env [temperature] of patch-here
  if not is-number? T-env [ set T-env optimal-temperature ]

  ;; scaling factors
  let size-factor (weight / M-max) ^ beta
  let temp-factor Q10 ^ ((T-env - optimal-temperature) / 10)
  let age-factor 1 + (1 - (age-num / max-age)) * 0.5

  ;; metabolic rate (per tick)
  let M-total Met-base * size-factor * temp-factor * age-factor
  set metabolism-rate M-total
  set total-metabolism total-metabolism + M-total

  ;; baseline reference metabolism
  let baseline-size-factor 1
  let baseline-age-factor 1.5
  let baseline-temp-factor 1

  let baseline-metabolism Met-base *
      baseline-size-factor *
      baseline-temp-factor *
      baseline-age-factor

  ;; efficiency multipliers
  let efficiency M-total / baseline-metabolism

  set swim-efficiency              efficiency
  set osmoregulation-efficiency    efficiency
  set lipid-catabolism-efficiency  efficiency
end

```
