# Landward Movement
```{r, include=FALSE}
current_chapter <- "11-Landward_Migration.Rmd"
```
```{r, child="_download_button.Rmd"}
```

## Overview

This function simulates landward movement for fish moving upstream through riverine and estuarine systems, where agents adjust movement based on local water velocity, energetic condition, and the relative difficulty of swimming in a given flow. Within this submodel, upstream movement emerges from the interaction between hydrodynamic forcing, body size, and energetic state, as agents navigate spatially variable flow fields where velocity and depth constrain swimming difficulty and energetic cost. These relationships are consistent with allometric scaling in aquatic organisms [@brown_toward_2004] and observed behavioral responses of fish to hydrodynamic conditions during movement [@gibson_selective_2001].

Movement is structured through a least-cost framework that is recalculated under changing hydrodynamic conditions, where agents respond locally to gradients in flow resistance and path difficulty rather than optimizing globally. This formulation allows movement pathways, travel time, and energy expenditure to emerge dynamically from repeated interactions between organismal capacity and the hydrodynamic environment.

## Purpose

The purpose of this submodel is to represent upstream movement under realistic spatial hydrodynamic resistance. Agents evaluate flow velocity, movement difficulty, body condition, and patch level travel costs before advancing along a least cost path toward a home location.

## Entities, State Variables, and Scales

### Spatial and Temporal Scales

**Spatial Unit:** Patch (3 m x 3 m resolution)\
**Temporal Unit:** 5 minute time steps (tick)

### Global Variables

| Variable | Definition |
|:----------------------------------:|:----------------------------------:|
| **max-seaward-velocity** | Maximum seaward directed velocity in the domain, used for normalization. |
| **max-landward-velocity** | Maximum landward directed velocity in the domain. |

### Patch Variables

| Variable | Definition |
|:----------------------------------:|:----------------------------------:|
| **velocity** | Depth averaged hydrodynamic velocity at the patch. |
| **depth** | Water depth at the patch. |
| **patch-terrain** | Indicates whether the patch is water or land. |
| **cost-to-home** | Patch level cost field producing a least cost corridor. |
| **visits-by-alewife** | Number of times the agent visits the patch. |
| **ticks-spent-alewife** | Time spent on the patch. |

### Agent Variables

| Variable | Definition |
|:----------------------------------:|:----------------------------------:|
| **weight** | Body mass of the fish. |
| **speed** | Current swimming speed. |
| **prev-speed** | Previous time step swimming speed. |
| **max-speed** | Maximum sustained speed. |
| **min-speed** | Minimum allowable speed. |
| **swim-efficiency** | Scaling factor for acceleration and deceleration. |
| **difficulty-factor** | Difficulty of movement under local velocity conditions. |
| **planned-path** | Sequence of patches in the least cost route. |
| **home-patch** | Destination patch for movement. |
| **trail** | Record of patches visited by the agent. |

## Process Overview and Scheduling

1.  Compute swimming difficulty from hydrodynamic velocity and body
    size.

2.  Calculate swimming speed using desired-speed, velocity influence,
    and acceleration limits.

3.  Evaluate the least cost field and construct a path to the home
    patch.

4.  Move along the path for a distance proportional to swimming speed.

5.  Record patch visitation and time spent.

## Design Concepts

**Basic Principles:** Hydrodynamic constraints, difficulty scaling, and
body condition influence agent movement.

**Emergence:** movement routes, temporal occupancy, and movement speed
emerge from repeated interactions of flow and difficulty.

**Objectives:** Agents attempt to reach the home patch but do not
optimize globally. Movement follows the least cost route at each step.

**Sensing:** Agents sense velocity and depth of the local and
neighboring patches.

**Observation:** Paths, timing, and residency can be monitored
chronologically.

## Initialization

|       Variable        |    Initial Value    |          Justification           |
|:----------------------:|:----------------------:|:----------------------:|
|     **velocity**      |    user-defined     |     Hydrodynamic input data.     |
|      **weight**       |  species-specific   | Required for difficulty scaling. |
|     **max-speed**     |  species-specific   |    Physiological constraint.     |
|     **min-speed**     |  species-specific   |     Prevents speed collapse.     |
|  **swim-efficiency**  | species or constant |   Controls acceleration rate.    |
| **difficulty-factor** |          1          |   Neutral starting difficulty.   |
|    **prev-speed**     |   starting speed    |     Smooth initial movement.     |

## Submodels

### Swimming Difficulty

Swimming difficulty reflects the hydrodynamic effort required for the
agent to move through its current patch. Velocity is normalized between
the maximum observed landward and seaward velocities:

$$
 V_{\text{normalized}} = \frac{V_{\text{patch}} - V_{\min}}{V_{\max} - V_{\min}}  \hspace{2cm} \text{(C.64)} 
$$

Where:

-   $V_{normalized}$ is the normalized patch velocity (dimensionless).
-   $V_{patch}$ is the depth-averaged along-channel water velocity at
    the patch (m s⁻¹).
-   $V_{min}$ is the minimum observed velocity (m s⁻¹).
-   $V_{max}$ is the maximum observed velocity (m s⁻¹).

This normalized velocity is scaled by relative body size
$M_{\text{agent}} / M_{\max}$ and raised to a scaling exponent $k$
[@brown_toward_2004; @gibson_selective_2001]:

$$
 Df_{\text{raw}} = 
\left(
\frac{V_{\text{normalized}}}{A_{\text{mass}} / M_{\max}}
\right)^k  \hspace{2cm} \text{(C.65)} 
$$

Where:

-   $Df_{raw}$ is the raw hydrodynamic difficulty prior to mapping
    (dimensionless).
-   $V_{normalized}$ is the normalized patch velocity (m s⁻¹).
-   $A_{mass}$ is an agent's body mass used for allometric and energetic
    scaling (g).
-   $M_{max}$ is the maximum body mass across the population (g).
-   $k$ is the energy scaling exponent (dimensionless).

Difficulty is then mapped onto the one to ten range:

$$
 D_f = 1 + 9 \cdot Df_{\text{raw}}  \hspace{2cm} \text{(C.66)} 
$$

Where:

-   $D_f$ is the swimming difficulty index (dimensionless).
-   $Df_{raw}$ is the raw hydrodynamic difficulty prior to mapping
    (dimensionless).

and constrained:

$$
 D_f \in [1, 10]  \hspace{2cm} \text{(C.67)} 
$$

Where:

-   $D_f$ is the mapped swimming difficulty index (dimensionless, range
    1–10).

Difficulty increases when:

-   $M_{\text{agent}}$ is small
-   $|V_{\text{patch}}|$ approaches its maximum value
-   flow opposes movement

### Swimming Speed

Swimming speed depends on energetic condition, hydrodynamic difficulty,
and the effect of local flow. The agent first computes an energy factor:

$$
 E_{\text{factor}} = \frac{E_{\text{agent}}}{100}  \hspace{2cm} \text{(C.68)} 
$$

Where:

-   $E_{factor}$ is the normalized energy level of the agent
    (dimensionless).
-   $E_{agent}$ is the total internal energy available to the agent
    (energy units).

Velocity impact from hydrodynamics is:

$$
 V_{\text{impact}} = k \cdot V_{\text{patch}} \cdot 300  \hspace{2cm} \text{(C.69)} 
$$

Where:

-   $V_{impact}$ is the flow contribution to desired swimming speed (m
    tick⁻¹).
-   $k$ is the energy scaling exponent used in velocity, difficulty, and
    osmoregulation terms (dimensionless).
-   $V_{patch}$ is the depth-Averaged along-channel water velocity at
    the patch (m s⁻¹).
-   $300$ is the timestep conversion from seconds to one tick (s/tick,
    e.g., 5 minutes).

The desired swimming speed is:

$$
 V_{\text{desired}} =
\frac{V_{\max} \cdot E_{\text{factor}}}{D_f}
+ V_{\text{impact}}  \hspace{2cm} \text{(C.70)} 
$$

Where:

-   $V_{desired}$ is the desired swimming speed before constraints (m
    tick⁻¹).
-   $V_{max}$ is the maximum observed velocity (m tick⁻¹).
-   $E_{factor}$ is the normalized energy level of the agent
    (dimensionless).
-   $D_f$ is the mapped swimming difficulty index (dimensionless).
-   $V_{impact}$ is the flow contribution to desired swimming speed (m
    tick⁻¹).

Speed is constrained between a minimum and maximum:

$$
 V_{\text{desired}} 
= \min(V_{\max}, \max(V_{\min}, V_{\text{desired}}))  \hspace{2cm} \text{(C.71)} 
$$

Where:

-   $V_{desired}$ is the desired swimming speed before constraints (m
    tick⁻¹).
-   $V_{max}$ is the maximum observed velocity (m tick⁻¹).
-   $V_{min}$ is the minimum observed velocity (m tick⁻¹).

Acceleration is smoothed using the previous speed:

$$
 \Delta V_{max} = 0.5 \cdot swim_{Eff}  \hspace{2cm} \text{(C.72)} 
$$

Where:

-   $\Delta V_{max}$ is the maximum allowable speed change per tick (m
    s⁻¹/tick).
-   $swim_{Eff}$ is the efficiency multiplier applied to swimming
    energetics (%).

If $|V_{\text{desired}} - V_{\text{prev}}| > \Delta V_{\max}$,where
$V_{prev}$ is the previous speed then speed moves gradually toward the
desired value. Otherwise speed is set directly.

### Path Based Movement

Landward movement uses a least cost routing strategy instead of direct
vector movement. The number of patches traversed in a tick is based on
swimming speed:

$$
 D_{travel} = \frac{A_{speed}}{3}  \hspace{2cm} \text{(C.73)} 
$$

Where:

-   $D_{travel}$ is the distance traveled in the current tick (patch
    units), where 1 patch = 3 m.
-   $A_{speed}$ is the actual swimming speed used for movement (m
    tick⁻¹).

Movement follows:

1.  compute cost-to-home
2.  generate a path using decreasing cost
3.  move through the first $\lceil \mathrm{travel\_distance} \rceil$
    patches
4.  record patch visits and time spent
5.  append patch-here to the trail if new location

### Travel Cost

Travel cost for a candidate patch incorporates velocity magnitude,
slope, and difficulty:

$$
 \text{cost}_{\text{raw}}
= |V_{\text{patch}}|
+ 0.5 \cdot \max(0, \Delta \text{depth})  \hspace{2cm} \text{(C.74)} 
$$

Where:

-   $cost_{raw}$ is the base travel cost including velocity and slope
    (dimensionless).
-   $V_{patch}$ is the depth-Averaged along-channel water velocity at
    the patch (m s⁻¹).
-   $\Delta d$ is the change in depth between patches (m).

Difficulty multiplies the base cost:

$$
 \text{cost}
= \text{cost}_{\text{raw}} \cdot D_f  \hspace{2cm} \text{(C.75)} 
$$

Where:

-   $cost$ is the difficulty-scaled travel cost per patch (energy
    units).
-   $cost_{raw}$ is the base travel cost including velocity and slope
    (dimensionless).
-   $D_f$ is the mapped swimming difficulty index (dimensionless).

This produces a cost-to-home field representing hydrodynamic resistance
(dimensionless), which is used for route selection.

### Swimming Energy

Swimming incurs a metabolic cost that scales with base swimming cost,
difficulty, and swim efficiency.

Let $\beta = 0.75$.\
Let the effective multiplier be:

$$
 M_{swim} = \frac{D_f}{swim_{Eff}}  \hspace{2cm} \text{(C.76)} 
$$

Where:

-   $M_{swim}$ is the swimming energy multiplier (dimensionless).
-   $D_f$ is the mapped swimming difficulty index (dimensionless).
-   $swim_{Eff}$ is the efficiency multiplier applied to swimming
    energetics (dimensionless).

Where lower efficiency results in higher energetic cost.

Energy expenditure is:

$$
 E_{\text{swim}} = swim_{base} \cdot M_{swim}^\beta  \hspace{2cm} \text{(C.77)} 
$$

Where:

-   $E_{swim}$ is the energy cost of swimming for the tick (energy
    units).
-   $swim_{base}$ is the baseline energetic cost of swimming ((energy
    units) patch⁻¹).
-   $M_{swim}$ is the swimming energy multiplier (dimensionless).
-   $\beta$ is the exponent controlling swimming cost scaling
    (dimensionless).

Total energy is reduced accordingly:

$$
 E_{\text{agent}} = \max(0, E_{\text{agent}} - E_{\text{swim}})  \hspace{2cm} \text{(C.78)} 
$$

Where:

-   $E_{agent}$ is the total internal energy available to the agent
    (energy units).
-   $E_{swim}$ is the energy cost of swimming for the tick (energy
    units).

Higher difficulty and lower swim efficiency increase metabolic cost.

## Netlogo Implementation

```{netlogo}
;; ===================================================================
;;  LANDWARD MOVEMENT + HYDRODYNAMIC SWIMMING SUBMODEL DECLARATIONS
;; ===================================================================

globals [

  ;; Velocity normalization bounds
  max-seaward-velocity
  max-landward-velocity
]

patches-own [

  ;; Environmental attributes
  velocity
  depth
  patch-terrain   ;; "water" or "land"

  ;; Least-cost routing fields
  cost-to-home
  cost-to-sea
  cost-to-prey

  ;; Tracking fish patch-level behavior
  visits-by-fish
  ticks-spent-fish
]

turtles-own [

  ;; Body condition
  weight

  ;; Swimming behavior
  speed
  prev-speed
  max-speed
  min-speed
  swim-efficiency

  ;; Hydrodynamic resistance
  difficulty-factor

  ;; Energetics
  E-swim
  swim-base

  ;; movement targets
  home-patch
  migration-patch

  ;; Movement memory
  planned-path
  trail
  previous-patch
  previous-x
  previous-y
]


;; ===================================================================
;;  SWIMMING DIFFICULTY (LANDWARD)
;; ===================================================================

to calculate-difficulty-landward
  
  let M-max max [weight] of breed      ;; largest fish in population
  let M-agent weight                   ;; focal individual
  let V-max max-seaward-velocity
  let V-min max-landward-velocity
  let k 0.75                           ;; scaling exponent

  ;; Normalize velocity
  let normalized-velocity (velocity - V-min) / (V-max - V-min)

  ;; Difficulty from normalized velocity + size ratio
  let Df-raw (normalized-velocity / (M-agent / M-max)) ^ k
  
  ;; Scale to range 1–10
  set difficulty-factor (1 + 9 * Df-raw)
  set difficulty-factor max list 1 (min list 10 difficulty-factor)
end


;; ===================================================================
;;  SWIMMING SPEED (LANDWARD)
;; ===================================================================

to calculate-swimming-speed-landward
  
  let k -0.75                                    ;; velocity influence
  let energy-factor (energy / 100)               ;; 0–1 scale
  let velocity-impact (k * velocity * 300)       ;; 5-min timestep scaling

  ;; Preliminary target speed
  let desired-speed 
    (max-speed * energy-factor / difficulty-factor)
      + velocity-impact

  ;; Clamp desired speed to biologically realistic bounds
  set desired-speed min list max-speed (max list min-speed desired-speed)

  ;; Smooth acceleration / deceleration
  let max-speed-change 0.5
  set max-speed-change (max-speed-change * swim-efficiency)

  if abs(desired-speed - prev-speed) > max-speed-change [
    if desired-speed > prev-speed [
      set speed prev-speed + max-speed-change
    ]
    if desired-speed < prev-speed [
      set speed prev-speed - max-speed-change
    ]
  ]

  if abs(desired-speed - prev-speed) <= max-speed-change [
    set speed desired-speed
  ]

  set prev-speed speed
end


;; ===================================================================
;;  LEAST-COST ROUTING TO HOME (LANDWARD)
;; ===================================================================

to calculate-cost-to-home [homing-patch]

  ;; initialize routing fields
  ask home-patch  [ set cost-to-home 0 ]
  ask patch-here  [ set cost-to-home 1e9 ]

  let frontier (list home-patch)
  let df difficulty-factor                 ;; hydrodynamic penalty

  while [not empty? frontier] [
    let current first frontier
    set frontier but-first frontier

    ask current [
      let current-cost cost-to-home
      
      ask neighbors4 with [patch-terrain = "water"] [
        
        let travel-cost compute-travel-cost current self df
        let new-cost current-cost + travel-cost

        if new-cost < cost-to-home [
          set cost-to-home new-cost
          set frontier lput self frontier
        ]
      ]
    ]
  ]
end


;; ===================================================================
;;  BUILD OPTIMAL PATH ENTIRELY FROM COST-FIELD
;; ===================================================================

to-report build-path [current-patch homing-patch]

  let path (list current-patch)
  let cur current-patch
  let max-steps 2000

  while [cur != homing-patch and max-steps > 0] [
    set max-steps max-steps - 1

    let candidates [neighbors4] of cur
    set candidates candidates with [
      patch-terrain = "water" and
      cost-to-home < [cost-to-home] of cur
    ]

    if not any? candidates [
      report path
    ]

    let next-patch min-one-of candidates [cost-to-home]
    set path lput next-patch path
    set cur next-patch
  ]

  report path
end


;; ===================================================================
;;  TRAVEL COST UNDER HYDRODYNAMIC RESISTANCE
;; ===================================================================

to-report compute-travel-cost [from-patch to-patch df]

  ;; velocity resistance
  let raw-vel [velocity] of to-patch
  let velocity-resistance 
      ifelse-value is-number? raw-vel [abs raw-vel] [0]

  ;; depth slope component
  let from-level [depth] of from-patch
  let to-level   [depth] of to-patch
  let slope 0

  if (is-number? from-level and is-number? to-level) [
    set slope max list 0 (to-level - from-level)
  ]

  ;; hydrodynamic cost + slope penalty
  let base-cost velocity-resistance + (0.5 * slope)

  ;; difficulty multiplier
  let df2 ifelse-value is-number? df [df] [1]

  report max list 0 (base-cost * df2)
end


;; ===================================================================
;;  LANDWARD MOVEMENT — PATH EXECUTION
;; ===================================================================

to migrate-landward
  
  let travel-distance (speed / 3)
  let path build-path patch-here home-patch
  set planned-path path
  
  if not empty? path [

    let steps-to-move ceiling travel-distance
    let move-patches sublist path 0 (min list steps-to-move length path)
    
    let old-patch patch-here

    foreach move-patches [ p ->

      ;; Record patch visitation
      if p != old-patch [
        ask p [ set visits-by-alewife visits-by-alewife + 1 ]
        set old-patch p
      ]

      ;; Move into patch
      face p
      move-to p

      ask p [
        set ticks-spent-alewife ticks-spent-alewife + 1
      ]
    ]
  ]

  ;; Add patch to historical trail
  if not member? patch-here trail [
    set trail lput patch-here trail
  ]
end


;; ===================================================================
;;  UNIVERSAL SWIMMING ENERGY COST
;; ===================================================================

to calculate-swim-energy
  
  let beta 0.75

  let energy-multiplier 
      (difficulty-factor * (1 / swim-efficiency))

  set E-swim 
      (swim-base * (energy-multiplier ^ beta))

  set energy max list 0 (energy - E-swim)
end
```
