# Seaward Movement
```{r, include=FALSE}
current_chapter <- "12-Seaward_Migration.Rmd"
```
```{r, child="_download_button.Rmd"}
```

## Overview

This function simulates seaward movement for fish moving downstream through riverine and estuarine systems. Agents respond to hydrodynamic conditions by adjusting swimming difficulty, swimming speed, and movement along a least cost pathway toward a downstream movement patch.

This submodel represents downstream movement following the same
mechanistic framework as landward movement, where movement emerges from
the interaction between hydrodynamic forcing, body size, and energetic
state [@brown_toward_2004; @gibson_selective_2001]. In this case,
movement occurs in the seaward direction, requiring a change in sign
within velocity-based scaling terms to reflect alignment with downstream
flow. As in landward movement, agents follow a least-cost pathway based
on local hydrodynamic resistance, allowing downstream trajectories,
travel time, and energy expenditure to emerge dynamically from the
interaction between organismal capacity and the flow field.

## Purpose

The purpose of this submodel is to represent downstream movement under realistic hydrodynamic resistance. Agents evaluate flow velocity, movement difficulty, energetic condition, and travel costs before advancing along a least cost route toward a movement destination.

## Entities, State Variables, and Scales

### Spatial and Temporal Scales

**Spatial Unit:** Patch (3 m x 3 m resolution)\
**Temporal Unit:** 5 minute time steps (tick)

### Global Variables

| Variable | Definition |
|:--:|:--:|
| **max-seaward-velocity** | Maximum observed seaward velocity in the domain. |
| **max-landward-velocity** | Maximum observed landward velocity in the domain. |

### Patch Variables

| Variable | Definition |
|:--:|:--:|
| **velocity** | Depth averaged hydrodynamic velocity. |
| **depth** | Water depth at the patch. |
| **patch-terrain** | Indicates whether the patch is land or water. |
| **cost-to-sea** | Patch-level travel cost for seaward movement. |
| **visits-by-alewife** | Number of times the agent visits the patch. |
| **ticks-spent-alewife** | Time spent on the patch. |

### Agent Variables

| Variable | Definition |
|:--:|:--:|
| **weight** | Body mass of the fish. |
| **speed** | Current swimming speed. |
| **prev-speed** | Previous time step swimming speed. |
| **max-speed** | Maximum sustained speed. |
| **min-speed** | Minimum allowable speed. |
| **swim-efficiency** | Swimming acceleration and energy scaling. |
| **difficulty-factor** | Hydrodynamic difficulty. |
| **planned-path** | Sequence of patches in least cost route. |
| **migration-patch** | Destination patch. |
| **trail** | Record of visited patches. |

## Process Overview and Scheduling

1.  Calculate swimming difficulty using normalized velocity and body
    size.
2.  Update swimming speed using energy reserves, difficulty, and flow
    velocity.
3.  Calculate cost-to-sea and construct least cost path.
4.  Move along the first part of the path based on swimming speed.
5.  Record patch visits and time spent.

## Design Concepts

**Basic Principles:** Hydrodynamic forcing, body size, and energy
availability determine downstream movement.

**Emergence:** Route choice, travel duration, and energetic depletion
emerge from repeated state updates.

**Objectives:** Agents attempt to reach the seaward migration patch
following a hydrodynamically efficient route.

**Sensing:** Agents sense local velocity, slope, difficulty, and patch
cost gradients.

**Observation:** Movement trajectories, energy decline, and convergence
routes can be monitored.

## Initialization

| Variable | Initial Value | Justification |
|:--:|:--:|:--:|
| **velocity** | Hydrodynamic input | Required environmental forcing. |
| **weight** | Species-specific | Required for difficulty scaling. |
| **max-speed** | Species-specific | Physiological constraint. |
| **min-speed** | Species-specific | Prevents collapse to zero speed. |
| **swim-efficiency** | Species or constant | Controls acceleration and deceleration. |
| **difficulty-factor** | 1 | Neutral start value. |
| **prev-speed** | Starting speed | Smooth initial movement. |

## Submodels

### Swimming Difficulty

Velocity is normalized between minimum and maximum observed values:

$$
 V_{normalized}
= 
\frac{V_{\text{patch}} - V_{\min}}
{V_{\max} - V_{\min}}  \hspace{2cm} \text{(C.79)} 
$$

Where:

-   $V_{normalized}$ is the normalized patch velocity (dimensionless).

-   $V_{patch}$ is the depth-averaged along-channel water velocity at
    the patch (m s⁻¹).

-   $V_{min}$ is the minimum observed velocity (m s⁻¹).

-   $V_{max}$ is the maximum observed velocity (m s⁻¹).

Difficulty scales with body size [@brown_toward_2004;
@gibson_selective_2001]:

$$
 Df_{\text{raw}}
=
\left(
\frac{V_{normalized}}
{A_{\text{mass}} / M_{\max}}
\right)^k  \hspace{2cm} \text{(C.80)} 
$$

Where:

-   $Df_{raw}$ is the raw hydrodynamic difficulty prior to mapping
    (dimensionless).

-   $V_{normalized}$ is the normalized patch velocity (dimensionless).

-   $A_{mass}$ is an agent's body mass used for allometric and energetic
    scaling (g).

-   $M_{max}$ is the maximum body mass across the population (g).

-   $k$ is the energy scaling exponent (dimensionless).

Difficulty is mapped onto the one-to-ten range:

$$
 D_f = 1 + 9 \cdot Df_{\text{raw}}  \hspace{2cm} \text{(C.81)} 
$$

Where:

-   $D_f$ is the mapped swimming difficulty index (dimensionless).
-   $Df_{raw}$ is the raw hydrodynamic difficulty prior to mapping
    (dimensionless).

and constrained:

$$
 D_f \in [1, 10]  \hspace{2cm} \text{(C.82)} 
$$

Where:

-   $D_f$ is the mapped swimming difficulty index (dimensionless, range
    1-10).

### Swimming Speed

Energy factor:

$$
 E_{\text{factor}} = \frac{E_{\text{agent}}}{100}  \hspace{2cm} \text{(C.83)} 
$$

Where:

-   $E_{factor}$ is the normalized energy level of the agent
    (dimensionless).

-   $E_{agent}$ is the total internal energy available to the agent
    (energy units).

Velocity impact from hydrodynamics is:

$$  V_{\text{impact}} = k \cdot V_{\text{patch}} \cdot 300  \hspace{2cm} \text{(C.69)}  $$

Where:

-   $V_{impact}$ is the flow contribution to desired swimming speed (m
    tick⁻¹).
-   $k$ is the energy scaling exponent used in velocity, difficulty, and
    osmoregulation terms (dimensionless).
-   $V_{patch}$ is the depth-Averaged along-channel water velocity at
    the patch (m s⁻¹).
-   $300$ is the timestep conversion from seconds to one tick (s/tick,
    e.g., 5 minutes).

The desired swimming speed is:

$$  V_{\text{desired}} = \frac{V_{\max} \cdot E_{\text{factor}}}{D_f} + V_{\text{impact}}  \hspace{2cm} \text{(C.70)}  $$

Where:

-   $V_{desired}$ is the desired swimming speed before constraints (m
    tick⁻¹).
-   $V_{max}$ is the maximum observed velocity (m tick⁻¹).
-   $E_{factor}$ is the normalized energy level of the agent
    (dimensionless).
-   $D_f$ is the mapped swimming difficulty index (dimensionless).
-   $V_{impact}$ is the flow contribution to desired swimming speed (m
    tick⁻¹).

Speed is constrained between a minimum and maximum:

$$  V_{\text{desired}}  = \min(V_{\max}, \max(V_{\min}, V_{\text{desired}}))  \hspace{2cm} \text{(C.71)}  $$

Where:

-   $V_{desired}$ is the desired swimming speed before constraints (m
    tick⁻¹).
-   $V_{max}$ is the maximum observed velocity (m tick⁻¹).
-   $V_{min}$ is the minimum observed velocity (m tick⁻¹).

Acceleration smoothing:

$$
 \Delta V_{\max} = 0.5 \cdot swim_{Eff}  \hspace{2cm} \text{(C.87)} 
$$

Where:

-   $\Delta V_{max}$ is the maximum allowable speed change per tick (m
    tick⁻¹).
-   $swim_{Eff}$ is the efficiency multiplier applied to swimming
    energetics (%).

If $|V_{\text{desired}} - V_{\text{prev}}| > \Delta V_{\max}$,where
$V_{prev}$ is the previous speed then speed moves gradually toward the
desired value. Otherwise speed is set directly.

### Path-Based Movement

Travel distance:

$$
 D_{travel}
=
\frac{A_{\text{speed}}}{3}  \hspace{2cm} \text{(C.88)} 
$$

Where:

-   $D_{travel}$ is the distance traveled in the current tick (patch
    units), where 1 patch = 3 m.
-   $A_{speed}$ is the actual swimming speed used for movement (m
    tick⁻¹).

Movement follows:

1.  compute cost-to-sea
2.  build path using decreasing cost
3.  traverse first $\lceil D_{travel} \rceil$ patches
4.  record visitation and time spent
5.  append to the movement trail

### Travel Cost

Raw cost:

$$
 \text{cost}_{\text{raw}}
=
|V_{\text{patch}}|
+
0.5 \cdot \max(0, \Delta d)  \hspace{2cm} \text{(C.89)} 
$$

Where:

-   $cost_{raw}$ is the base travel cost including velocity and slope
    (dimensionless).
-   $V_{patch}$ is the depth-averaged along-channel water velocity at
    the patch (m s⁻¹).
-   $\Delta d$ is the change in depth between patches (m).

Difficulty scaling:

$$
 \text{cost}
=
\text{cost}_{\text{raw}} \cdot D_f  \hspace{2cm} \text{(C.90)} 
$$

Where:

-   $cost$ is the difficulty-scaled travel cost per patch (energy
    units).

-   $cost_{raw}$ is the base travel cost including velocity and slope
    (dimensionless).

-   $D_f$ is the mapped swimming difficulty index (dimensionless).

This produces a cost-to-sea field representing hydrodynamic resistance
(dimensionless), which is used for route selection.

### Swimming Energy

Let $\beta = 0.75$.

Energy multiplier:

$$
 M_{swim} = \frac{D_f}{swim_{Eff}}  \hspace{2cm} \text{(C.91)} 
$$

Where:

-   $M_{swim}$ is the swimming energy multiplier (dimensionless).
-   $D_f$ is the mapped swimming difficulty index (dimensionless).
-   $swim_{Eff}$ is the efficiency multiplier applied to swimming
    energetics (%).

Energy cost:

$$
 E_{\text{swim}}
=
swim_{base} \cdot M_{swim}^{\beta}  \hspace{2cm} \text{(C.92)} 
$$

Where:

-   $E_{swim}$ is the energy cost of swimming for the tick (energy
    units).

-   $swim_{base}$ is the baseline energetic cost of swimming ((energy
    units) patch⁻¹).

-   $M_{swim}$ is the swimming energy multiplier (dimensionless).

-   $\beta$ is the exponent controlling swimming cost scaling
    (dimensionless).

Energy update:

$$
 E_{\text{agent}}
=
\max \left(0,\, E_{\text{agent}} - E_{\text{swim}} \right)  \hspace{2cm} \text{(C.93)} 
$$

Where:

-   $E_{factor}$ is the normalized energy level of the agent (energy
    units).
-   $E_{agent}$ is the total internal energy available to the agent
    (energy units).

## NetLogo Implementation

```{netlogo}

;; ===================================================================
;;  SEAWARD MOVEMENT + HYDRODYNAMIC SWIMMING SUBMODEL DECLARATIONS
;; ===================================================================

globals [

  ;; Velocity normalization bounds
  max-seaward-velocity
  max-landward-velocity
]

patches-own [

  ;; Environmental attributes
  velocity
  depth
  patch-terrain   ;; "water" or "land"

  ;; Least-cost routing fields
  cost-to-home
  cost-to-sea
  cost-to-prey

  ;; Tracking fish patch-level behavior
  visits-by-fish
  ticks-spent-fish
]

turtles-own [

  ;; Body condition
  weight

  ;; Swimming behavior
  speed
  prev-speed
  max-speed
  min-speed
  swim-efficiency

  ;; Hydrodynamic resistance
  difficulty-factor

  ;; Energetics
  E-swim
  swim-base

  ;; Movement targets
  home-patch
  migration-patch

  ;; Movement memory
  planned-path
  trail
  previous-patch
  previous-x
  previous-y
]

;; ===================================================================
;;  SEAWARD MOVEMENT + HYDRODYNAMIC SWIMMING SUBMODEL
;; ===================================================================

;; -------------------------------------------------------------------
;;  Swimming Difficulty (Seaward)
;; -------------------------------------------------------------------
to calculate-difficulty-seaward
  
  let M-max max [weight] of breed
  let M-agent weight
  let V-max max-landward-velocity
  let V-min max-seaward-velocity
  let k 0.75

  let normalized-velocity (velocity - V-min) / (V-max - V-min)

  let Df-raw (normalized-velocity / (M-agent / M-max)) ^ k

  set difficulty-factor (1 + 9 * Df-raw)
  set difficulty-factor max list 1 (min list 10 difficulty-factor)
end


;; -------------------------------------------------------------------
;;  Swimming Speed (Seaward)
;; -------------------------------------------------------------------
to calculate-swimming-speed-seaward
  
  let k 0.75
  let energy-factor (energy / 100)

  let velocity-impact (k * velocity * 300)

  let desired-speed 
    (max-speed * energy-factor / difficulty-factor)
      + velocity-impact

  set desired-speed 
      min list max-speed (max list min-speed desired-speed)

  let max-speed-change 0.5
  set max-speed-change (max-speed-change * swim-efficiency)

  if abs(desired-speed - prev-speed) > max-speed-change [
    if desired-speed > prev-speed [
      set speed prev-speed + max-speed-change
    ]
    if desired-speed < prev-speed [
      set speed prev-speed - max-speed-change
    ]
  ]

  if abs(desired-speed - prev-speed) <= max-speed-change [
    set speed desired-speed
  ]

  set prev-speed speed
end


;; -------------------------------------------------------------------
;;  Cost-to-Sea (Least Cost Routing)
;; -------------------------------------------------------------------
to calculate-cost-to-sea [sea-patch]

  ask migration-patch [ set cost-to-sea 0 ]
  ask patch-here     [ set cost-to-sea 1e9 ]

  let frontier (list migration-patch)
  let df difficulty-factor

  while [not empty? frontier] [

    let current first frontier
    set frontier but-first frontier

    ask current [

      let current-cost cost-to-sea

      ask neighbors4 with [patch-terrain = "water"] [

        let travel-cost compute-travel-cost current self df
        let new-cost current-cost + travel-cost

        if new-cost < cost-to-sea [
          set cost-to-sea new-cost
          set frontier lput self frontier
        ]
      ]
    ]
  ]
end


;; -------------------------------------------------------------------
;;  Build Path to Sea
;; -------------------------------------------------------------------
to-report build-path-sea [current-patch sea-patch]

  let path (list current-patch)
  let cur current-patch
  let max-steps 2000

  while [cur != migration-patch and max-steps > 0] [

    set max-steps max-steps - 1

    let candidates [neighbors] of cur
    set candidates candidates with [
      patch-terrain = "water" and 
      cost-to-sea < [cost-to-sea] of cur
    ]

    if not any? candidates [
      report path
    ]

    let next-patch min-one-of candidates [cost-to-sea]
    set path lput next-patch path
    set cur next-patch
  ]

  report path
end


;; -------------------------------------------------------------------
;;  Seaward Movement – Movement Execution
;; -------------------------------------------------------------------
to migrate-seaward
  
  ;; Schooling logic (excluded in documentation)
  find-schoolmates
  if any? schoolmates [
    find-nearest-neighbor
    ifelse distance nearest-neighbor < minimum-separation [
      separate
    ][
      cohere
      align
    ]
    adjust-speed
  ]

  ;; Path-based movement
  let travel-distance (speed / 3)
  let path build-path-sea patch-here migration-patch
  set planned-path path

  if not empty? path [

    let steps-to-move ceiling travel-distance
    let move-patches sublist path 0 (min list steps-to-move length path)

    let old-patch patch-here

    foreach move-patches [ p ->

      if p != old-patch [
        ask p [ set visits-by-alewife visits-by-alewife + 1 ]
        set old-patch p
      ]

      face p
      move-to p

      ask p [
        set ticks-spent-alewife ticks-spent-alewife + 1
      ]
    ]
  ]

  if not member? patch-here trail [
    set trail lput patch-here trail
  ]
end


;; -------------------------------------------------------------------
;;  Universal Swimming Energy Cost
;; -------------------------------------------------------------------
to calculate-swim-energy
  
  let beta 0.75
  let energy-multiplier (difficulty-factor * (1 / swim-efficiency))

  set E-swim (swim-base * (energy-multiplier ^ beta))
  set energy max list 0 (energy - E-swim)
end
```

