# Selective Tidal Stream Transport
```{r, include=FALSE}
current_chapter <- "14-Selective_Tidal_Stream_Transport.Rmd"
```
```{r, child="_download_button.Rmd"}
```

## Overview

Selective Tidal Stream Transport (STST) is a behavioral strategy that enables agents to conserve energy by alternating between active swimming and passive drift under strong hydrodynamic forcing. Within this submodel, STST is triggered when the along-channel velocity of a patch exceeds an agent’s effective swimming speed and that speed falls below a species-specific minimum threshold. When engaged, agents align with the tidal current and are passively transported upstream or downstream depending on flow direction, reducing the metabolic cost of movement by substituting active swimming with drift.

This behavior is well documented in anadromous and estuarine fishes, where individuals exploit tidal currents to facilitate migration while minimizing energy expenditure [@ForwardTankersley2001; @Gibson2001STST]. Movement during STST is governed by local flow velocity, allowing migration pathways, displacement, and energy savings to emerge dynamically from the interaction between individual swimming capacity and the tidal flow field. STST persists for a limited duration or until swimming ability improves, after which agents resume directional migration.

## Purpose

To simulate a passive energy-conserving behavior in migratory fish that allows them to use tidal currents to move when swimming capacity is insufficient to overcome flow velocities.

## Entities, State Variables, and Scales

### Spatial and Temporal Scales

**Spatial Unit:** Patch (3 m x 3 m resolution)\
**Temporal Unit:** 5 minute time steps (tick)

### **Global Variables**

| Variable Name | Definition |
|----|----|
| **max-seaward-velocity** | Highest magnitude seaward-directed hydrodynamic velocity across the spatial domain. |
| **max-landward-velocity** | Highest magnitude landward-directed hydrodynamic velocity across the spatial domain. |

### Patch Variables

| Variable Name | Definition |
|:----------------------------------:|:----------------------------------:|
| **Velocity** $V_{patch}$ | The along-channel velocity of a given patch, derived from hydrodynamic model inputs, where positive values are in the landward direction and negative values are in the seaward direction. |
| **tidal-transport-in-patch** | Count of agents exhibiting tidal stream transport within a patch (for habitat quality analysis). |

### Agent Variables

|  |  |
|:----------------------------------:|:----------------------------------:|
| **Variable Name** | **Definition** |
| **energy** $E_{agent}$ | Total energy available to the agent. |
| **swimming energy** $E_{swim}$ | Energy expenditure from movement per time step. |
| **base swim energy** $swim_{base}$ | Baseline energy cost of movement. |
| **swimming difficulty** $D_f$ | Velocity-based proxy representing hydrodynamic resistance. |
| **in-STST?** $STST_{?}$ | Boolean value indicating if the agent is actively in STST. |
| **swimming speed** $V_{agent}$ | The effective swimming speed of the agent. |
| **minimum threshold speed** $Speed_{min}$ | The minimum speed at which an agent will move. |

## Process Overview and Scheduling

1.  Compare swimming speed ($A_{speed}$) with flow speed ($V_{patch}$).

2.  If $|V_{patch}| > A_{speed}$ and $A_{speed} \leq Speed_{min}$, enter
    STST.

3.  In STST: align with current, update position via drift, apply
    reduced energy cost.

4.  If $A_{speed} > Speed_{min}$, exit STST and resume active swimming.

## Design Concepts

**Basic Principles:** Selective tidal stream transport (STST) is a
documented behavioral strategy in anadromous fishes that exploits tidal
currents to conserve energy during migration (e.g.,
@ForwardTankersley2001; @Gibson2001STST)

**Emergence:** Passive drift behavior and resulting migration paths
emerge from agent-flow interactions and individual swimming limitations.

**Adaptation:** Agents adapt their mode of movement based on their
swimming ability relative to environmental flow, dynamically choosing
energy-efficient strategies.

**Objectives:** Agents seek to minimizing energy loss in strong flows.

**Sensing:** Agents sense their own $A_{speed}$ and the $V_{patch}$ to
determine whether passive drift is needed.

**Observation:** Records STST patch events , energy expenditure
($E_{agent}$), and displacement are logged to analyze behavior across
flow regimes.

## Initialization

| Variable | Initialized Value | Justification |
|:----------------------:|:----------------------:|:----------------------:|
| $V_{agent}$ | Based on size, energy, and difficulty factor | Reflects agent's swimming capability based on metabolic limits. |
| $swim_{min}$ | Species-specific parameter | Represents the minimum sustained swimming velocity of the agent. |
| $swim_{max}$ | Species-specific parameter | Represents the maximum sustained swimming velocity of the agent. |
| $E_{agent}$ | 100 | Assumes full energy at the start of simulation or at spawning. |
| $swim_{base}$ | $0.02 \cdot \frac{M_{agent}}{M_{max}}^{k}$ | Scales locomotion cost nonlinearly with size; can be calibrated. |

## Submodels

### Trigger Conditions for STST

Agents compare their swimming ability to the flow conditions. If local
flow exceeds their capability and their effort is below a defined
threshold, they enter STST:

$$
 |V_{patch}| > A_{speed} \quad \text{and} \quad A_{speed} \leq Speed_{min}  \hspace{2cm} \text{(C.94)} 
$$

Where:

-   $V_{patch}$ is the depth-Averaged along-channel water velocity at
    the patch (m s⁻¹).
-   $A_{speed}$ is the actual swimming speed used for movement (m s⁻¹).
-   $Speed_{min}$ is the minimum sustainable swimming speed (m s⁻¹).

While in STST:

Heading aligns with the current (drift vector) & swimming speed is set
to:

$$
 A_{speed} = |V_{patch}|  \hspace{2cm} \text{(C.95)} 
$$

Where:

-   $V_{patch}$ is the depth-Averaged along-channel water velocity at
    the patch (m s⁻¹).
-   $A_{speed}$ is the actual swimming speed used for movement (m s⁻¹).

Energy is set as:

$$
 E_{swim} = swim_{base}  \hspace{2cm} \text{(C.96)} 
$$

Where:

-   $E_{swim}$ is the energy cost of swimming for the tick (energy
    units).
-   $swim_{base}$ is the baseline energetic cost of swimming (energy
    units).

### Behavior During STST

While in STST, agents align with the current (either landward or
seaward) and are passively transported:

$$
 \vec{Y}_{t+1} = \vec{Y}_t + |V_{patch}| \cdot \hat{u}  \hspace{2cm} \text{(C.97)} 
$$

Where:

-   $\vec{Y}_t$ is the Agent spatial position vector.
-   $V_{patch}$ is the depth-Averaged along-channel water velocity at
    the patch (m s⁻¹).
-   $\hat{u}_{patch}$ is the unit direction vector of patch flow,
    defined as $\hat{u}_{patch} = V_{patch} / |V_{patch}|$ when
    $|V_{patch}| > 0$.

Swimming speed is overwritten:

$$
 A_{speed} = |V_{patch}|  \hspace{2cm} \text{(C.98)} 
$$

Where:

-   $A_{speed}$ is the actual swimming speed used for movement (m s⁻¹).
-   $V_{patch}$ is the depth-Averaged along-channel water velocity at
    the patch (m s⁻¹).

Energy cost is minimized:

$$
 E_{swim} = swim_{base}  \hspace{2cm} \text{(C.99)} 
$$

Where:

-   $E_{swim}$ is the energy cost of swimming for the tick (energy
    units).
-   $swim_{base}$ is the baseline energetic cost of swimming (energy
    units).
    
### Stop Conditions for STST

Agents exit STST when they regain sufficient swimming capacity to exceed
threshold:

$$
 A_{speed} > speed_{min}  \hspace{2cm} \text{(C.100)} 
$$

Where:

-   $A_{speed}$ is the actual swimming speed used for movement (m s⁻¹).
-   $speed_{min}$ is the minimum sustainable swimming speed (m s⁻¹).

This triggers a return to active migratory movement and deactivates
$STST_{?}$.

## Netlogo Implementation

```{netlogo}
;; -------------------------------------------------------------------
;;  GLOBALS
;; -------------------------------------------------------------------
globals [
  max-seaward-velocity
  max-landward-velocity
  
  landward-migration?
  seaward-migration?

  selective-tidal-transport?
  STST-start-tick
]

;; -------------------------------------------------------------------
;;  PATCH VARIABLES
;; -------------------------------------------------------------------
patches-own [
  velocity
  depth
  patch-terrain
  
  tidal-transport-in-patch
]

;; -------------------------------------------------------------------
;;  TURTLE VARIABLES
;; -------------------------------------------------------------------
turtles-own [
  speed
  min-speed
  swim-efficiency
  swim-base
  difficulty-factor
  energy
  E-swim
  
  selective-tidal-transport?
  STST-start-tick
]

;; -------------------------------------------------------------------
;;  STST — LANDWARD
;; -------------------------------------------------------------------
to selective-tidal-stream-transport-landward
  let V-patch [velocity] of patch-here
  let S-swim (speed / 300) ;; 5 minute speed

  ;; Continue drifting (1 tick cooldown)
  if selective-tidal-transport? and (ticks - STST-start-tick < 1) [
    let drift-distance abs V-patch * (1 - swim-efficiency)
    let drift-target one-of neighbors with [patch-terrain = "water"]
    if drift-target != nobody [ safe-move drift-target drift-distance ]
    set tidal-transport-in-patch tidal-transport-in-patch + 1
    stop
  ]

  ;; Enter STST
  if abs V-patch > S-swim and S-swim <= min-speed [
    set selective-tidal-transport? true
    set STST-start-tick ticks
    let drift-distance abs V-patch * (1 - swim-efficiency)
    let drift-target one-of neighbors with [patch-terrain = "water"]
    if drift-target != nobody [ safe-move drift-target drift-distance ]
    set tidal-transport-in-patch tidal-transport-in-patch + 1
    stop
  ]

  ;; Exit STST → resume active migration
  if S-swim > min-speed or (ticks - STST-start-tick >= 1) [
    set selective-tidal-transport? false
    calculate-swim-energy
    set energy max list 0 (energy - E-swim)
    migrate-landward
  ]
end

;; -------------------------------------------------------------------
;;  STST — SEAWARD
;; -------------------------------------------------------------------
to selective-tidal-stream-transport-seaward
  let V-patch [velocity] of patch-here
  let S-swim (speed / 300)

  ;; Continue STST drift
  if selective-tidal-transport? and (ticks - STST-start-tick < 1) [
    let drift-distance abs V-patch * (1 - swim-efficiency)
    let drift-target one-of neighbors with [patch-terrain = "water"]
    if drift-target != nobody [ safe-move drift-target drift-distance ]
    set tidal-transport-in-patch tidal-transport-in-patch + 1
    stop
  ]

  ;; Enter STST
  if abs V-patch > S-swim and S-swim <= min-speed [
    set selective-tidal-transport? true
    set STST-start-tick ticks
    let drift-distance abs V-patch * (1 - swim-efficiency)
    let drift-target one-of neighbors with [patch-terrain = "water"]
    if drift-target != nobody [ safe-move drift-target drift-distance ]
    set tidal-transport-in-patch tidal-transport-in-patch + 1
    stop
  ]

  ;; Exit STST → resume seaward migration
  if S-swim > min-speed or (ticks - STST-start-tick >= 1) [
    set selective-tidal-transport? false
    calculate-swim-energy
    set energy max list 0 (energy - E-swim)
    migrate-seaward
  ]
end

;; -------------------------------------------------------------------
;;  SAFE MOVEMENT HANDLER
;; -------------------------------------------------------------------
to safe-move [target-patch move-speed]
  face target-patch
  let next-patch patch-ahead move-speed

  ifelse [patch-terrain] of next-patch = "water" [
    fd move-speed
  ] [
    let nearby-water one-of patches in-radius 1 with [patch-terrain = "water"]
    if nearby-water != nobody [ move-to nearby-water ]
  ]
end
```
