# Simple Model Building Tutorial: Schooling & Landward Movement
```{r, include=FALSE}
current_chapter <- "16-integrate_functions_example_01.Rmd"
```
```{r, child="_download_button.Rmd"}
```

This tutorial introduces function integration by combining two submodels — **schooling** and **landward movement** — into a single model time step. It is intentionally simple: movement is represented as a generic forward step (`fd speed`) with no hydrodynamic resistance, routing cost fields, or energy tracking. The goal is to show *how* these two functions talk to each other and in what order they should be called, before adding the complexity of real hydrology or site-specific data.

## Module Integration

Schooling and landward movement interact because they share the same agents and operate within the same time step, but serve different purposes. movement drives each agent toward an upstream destination by turning to face `home-patch` and stepping forward; schooling adjusts heading and speed so that agents moving in the same direction stay cohesive as a group. When both functions are active together, movement sets the gross direction of travel and schooling refines it. Within this simple example, agents move toward their destination while also maintaining spacing and alignment with nearby con-specifics.

**Temporal structure.** Each tick represents one discrete time step. Every agent does the following, in order:

1.  Check whether movement has started
2.  If migrating: face `home-patch` and move forward by `speed`
3.  If schoolmates are present: apply separation, cohesion, and alignment rules
4.  Check arrival at destination

**Spatial scale.** The environment is a simple abstract grid of water patches. There are no real coordinates, tidal cycles, or input data files. Agents have a `home-patch` (upstream destination) and a `migration-patch` (downstream entry point), which are set at initialization.

**Coupling logic.** movement faces the agent toward `home-patch` and calls `fd speed`; schooling then adjusts heading and speed *after* movement, modifying the next time step's trajectory. This ordering means movement drives direction and schooling refines it. Because `adjust-speed` can slow an agent below its set speed, the school naturally converges on an emergent group speed — this is intentional behavior representing the cohesion trade-off.

## Function Dependencies

| Function | Required Inputs | Output Variables | Dependent On |
|------------------|------------------|------------------|------------------|
| `migrate-landward` | `speed`, `home-patch` | `xcor`, `ycor`, `patch-here` | movement state flags |
| `find-schoolmates` | `breed`, `start-migration?`, `landward-migration?` | `schoolmates` | movement state flags |
| `find-nearest-neighbor` | `schoolmates` | `nearest-neighbor` | `find-schoolmates` |
| `separate` | `nearest-neighbor`, `minimum-separation`, `speed`, `min-speed` | `heading`, `speed` | `find-nearest-neighbor` |
| `cohere` | `schoolmates`, `cohere-coefficient` | `heading` | `find-nearest-neighbor` |
| `align` | `schoolmates`, `align-coefficient` | `heading` | `find-nearest-neighbor` |
| `adjust-speed` | `schoolmates`, `speed`, `min-speed` | `speed` | `find-nearest-neighbor` |

**Key dependency:** Schooling requires `start-migration?` and `landward-migration?` to be `true` before `find-schoolmates` will return any agents. An agent that has not yet begun migrating will have an empty `schoolmates` agent set and skip the schooling block entirely.

## Implementation in NetLogo

This example uses a minimal environment — all patches are water and agents face and step toward `home-patch` each time step. There are no GIS data, no CSV inputs, no calendar, and no energy or velocity calculations. A complete working implementation is available at [Simple Model NetLogo Implementation](https://github.com/vmahan1998/Sample_ABM_Tutorials/tree/main/Simple_Example_01/Netlogo).

### Variable declarations

``` netlogo
patches-own [
  patch-terrain   ;; "water" or "land"
]

turtles-own [
  ;; Migration state
  start-migration?
  landward-migration?
  seaward-migration?
  at-destination?

  ;; Destination patches
  home-patch
  migration-patch

  ;; Swimming
  speed
  min-speed

  ;; Schooling
  schoolmates
  nearest-neighbor
  minimum-separation
  align-coefficient
  cohere-coefficient
  separate-coefficient
]
```

### Setup

``` netlogo
to setup
  clear-all

  ask patches [
    set patch-terrain "water"
  ]

  create-turtles 20 [
    setxy (min-pxcor + random world-width) (min-pycor + random world-height)
    set shape "fish"
    set color cyan

    ;; Migration state — start migrating immediately in this minimal example
    set start-migration?    true
    set landward-migration? true
    set seaward-migration?  false
    set at-destination?     false

    ;; Assign destination and entry patches
    set home-patch      patch max-pxcor 0   ;; upstream end
    set migration-patch patch min-pxcor 0   ;; downstream entry point

    ;; Swimming parameters
    set min-speed  0.5
    set speed      1 + random-float 1

    ;; Schooling parameters (BOIDS coefficients)
    set minimum-separation   1.5
    set align-coefficient    25
    set cohere-coefficient   25
    set separate-coefficient 25
  ]

  reset-ticks
end
```

### The go procedure — time sequence

``` netlogo
to go

  ask turtles [

    ;; Step 1: move toward home-patch
    ;; face turns the agent toward its destination; fd moves it forward by speed.
    ;; No routing cost field or velocity calculation needed here.
    if landward-migration? and not at-destination? [
      face home-patch
      fd speed
    ]

    ;; Step 2: apply schooling rules (separation, cohesion, alignment, speed)
    ;; find-schoolmates uses start-migration? and landward-migration? to form
    ;; groups only among agents sharing the same movement state.
    school

    ;; Step 3: check arrival
    if patch-here = home-patch [
      set at-destination?     true
      set landward-migration? false
    ]
  ]

  tick
end
```

### What to observe

-   **School cohesion:** Plot `mean [distance nearest-neighbor] of turtles with [any? schoolmates]` over time. It should stabilize near `minimum-separation`.
-   **Movement progress:** Plot `mean [xcor] of turtles` — it should increase steadily toward `max-pxcor`.
-   **Speed convergence:** Plot `mean [speed] of turtles` — because `adjust-speed` pulls agents toward the group speed, individual variation should narrow over time.

## Implementation in R

The R version represents the same logic using a data frame of agents updated in a loop. Movement is a simple position increment toward `home_x` — no routing, no velocity, no energy — and schooling adjusts heading and speed using the same BOIDS rules. A complete working implementation is available at [Simple Model R Implementation](https://github.com/vmahan1998/Sample_ABM_Tutorials/tree/main/Simple_Example_01/R).

### Agent data structure

``` r
library(dplyr)

n_agents <- 20
n_ticks  <- 150

agents <- tibble(
  id                   = 1:n_agents,
  x                    = runif(n_agents, 0, 5),   # random start positions
  y                    = runif(n_agents, -5, 5),
  home_x               = 50,                      # upstream destination
  start_migration      = TRUE,
  landward_migration   = TRUE,
  at_destination       = FALSE,
  speed                = runif(n_agents, 1, 2),
  min_speed            = 0.5,
  heading              = 0,                       # degrees, 0 = facing upstream
  minimum_separation   = 1.5,
  cohere_coefficient   = 0.25,
  align_coefficient    = 0.25,
  separate_coefficient = 0.25
)
```

### Submodel functions

``` r
# ── Landward Movement ────────────────────────────────────────────────────────
# Turns agent toward home_x and steps forward by speed.
# No velocity, cost field, or energy — just direction and distance.
migrate_landward <- function(x, y, speed, home_x) {
  angle <- atan2(0, home_x - x)        # heading toward home (y held at 0)
  new_x <- x + speed * cos(angle)
  new_x <- pmin(new_x, home_x)         # cap at destination
  list(x = new_x, y = y, heading = angle * 180 / pi)
}

# ── Find schoolmates ──────────────────────────────────────────────────────────
find_schoolmates <- function(agent_id, agents) {
  focal <- agents[agents$id == agent_id, ]
  agents[agents$id != agent_id &
           agents$start_migration &
           agents$landward_migration == focal$landward_migration, ]
}

# ── Schooling (BOIDS rules) ───────────────────────────────────────────────────
school_agent <- function(agent_id, agents) {
  focal <- agents[agents$id == agent_id, ]
  mates <- find_schoolmates(agent_id, agents)

  if (nrow(mates) == 0) {
    return(list(heading = focal$heading, speed = focal$speed))
  }

  # Nearest neighbor
  dists   <- sqrt((mates$x - focal$x)^2 + (mates$y - focal$y)^2)
  nn_dist <- min(dists)
  nn_x    <- mates$x[which.min(dists)]

  new_heading <- focal$heading
  new_speed   <- focal$speed

  if (nn_dist < focal$minimum_separation) {
    # Separate: slow down if neighbor is ahead, turn away otherwise
    if (nn_x > focal$x) {
      new_speed <- max(focal$min_speed, focal$speed - 0.1)
    } else {
      new_heading <- focal$heading +
        (focal$separate_coefficient / 100) * sign(focal$x - nn_x) * 10
    }
  } else {
    # Cohere: turn toward centroid of schoolmates
    centroid_x  <- mean(mates$x)
    new_heading <- focal$heading +
      (focal$cohere_coefficient / 100) * sign(centroid_x - focal$x) * 10

    # Align: match mean heading of schoolmates
    mean_heading <- mean(mates$heading)
    new_heading  <- new_heading +
      (focal$align_coefficient / 100) * (mean_heading - new_heading)
  }

  # Adjust speed: speed up if a schoolmate is faster, otherwise slow down
  if (max(mates$speed) > focal$speed + 0.1) {
    new_speed <- focal$speed * 1.1
  } else if (focal$speed > focal$min_speed) {
    new_speed <- focal$speed * 0.9
  }
  new_speed <- max(new_speed, focal$min_speed)

  list(heading = new_heading %% 360, speed = new_speed)
}
```

### Simulation loop

``` r
log_ticks <- vector("list", n_ticks)

for (tick in 1:n_ticks) {

  # Step 1: migrate — face home and step forward
  migrating <- agents$start_migration & agents$landward_migration &
               !agents$at_destination

  move_results <- lapply(which(migrating), function(i) {
    migrate_landward(agents$x[i], agents$y[i],
                     agents$speed[i], agents$home_x[i])
  })

  agents$x[migrating]       <- sapply(move_results, `[[`, "x")
  agents$y[migrating]       <- sapply(move_results, `[[`, "y")
  agents$heading[migrating] <- sapply(move_results, `[[`, "heading")

  # Step 2: school — separation, cohesion, alignment, speed adjustment
  school_results <- lapply(agents$id, function(aid) {
    school_agent(aid, agents)
  })
  agents$heading <- sapply(school_results, `[[`, "heading")
  agents$speed   <- sapply(school_results, `[[`, "speed")

  # Step 3: check arrival
  agents <- agents |>
    mutate(
      at_destination     = x >= home_x,
      landward_migration = !at_destination
    )

  # Log summary statistics
  log_ticks[[tick]] <- tibble(
    tick       = tick,
    mean_x     = mean(agents$x),
    mean_speed = mean(agents$speed),
    n_arrived  = sum(agents$at_destination)
  )
}

results <- bind_rows(log_ticks)
```

### Visualizing outputs

``` r
library(ggplot2)

# Migration progress
ggplot(results, aes(x = tick, y = mean_x)) +
  geom_line(color = "#2c7bb6", linewidth = 1) +
  labs(title = "Mean agent position over time",
       x = "Tick", y = "Mean upstream position") +
  theme_minimal()

# Speed convergence from schooling
ggplot(results, aes(x = tick, y = mean_speed)) +
  geom_line(color = "#d7191c", linewidth = 1) +
  labs(title = "Mean agent speed over time",
       x = "Tick", y = "Mean speed") +
  theme_minimal()

# Cumulative arrivals
ggplot(results, aes(x = tick, y = n_arrived)) +
  geom_step(color = "#1a9641", linewidth = 1) +
  labs(title = "Cumulative arrivals at home patch",
       x = "Tick", y = "Number of agents arrived") +
  theme_minimal()
```

## Implementation in Python

The Python version follows the same structure as the R implementation — a data frame of agents updated in a loop, with movement as a simple step toward `home_x` and schooling applied per agent. The `pandas` library handles the agent table; schooling is applied with a row-wise loop since each agent needs to read the full table. A complete working implementation is available at [Simple Model Python Implementation](https://github.com/vmahan1998/Sample_ABM_Tutorials/tree/main/Simple_Example_01/Python).

### Agent data structure

``` python
import numpy as np
import pandas as pd

n_agents = 20
n_ticks  = 150
rng      = np.random.default_rng(seed=42)

agents = pd.DataFrame({
    "id"                  : np.arange(n_agents),
    "x"                   : rng.uniform(0, 5, n_agents),   # random start positions
    "y"                   : rng.uniform(-5, 5, n_agents),
    "home_x"              : 50,                            # upstream destination
    "start_migration"     : True,
    "landward_migration"  : True,
    "at_destination"      : False,
    "speed"               : rng.uniform(1, 2, n_agents),
    "min_speed"           : 0.5,
    "heading"             : 0.0,                           # degrees, 0 = upstream
    "minimum_separation"  : 1.5,
    "cohere_coefficient"  : 0.25,
    "align_coefficient"   : 0.25,
    "separate_coefficient": 0.25,
})
```

### Submodel functions

``` python
# ── Landward Movement ────────────────────────────────────────────────────────
# Turns agent toward home_x and steps forward by speed.
# No velocity, cost field, or energy — just direction and distance.

def migrate_landward(x, y, speed, home_x):
    angle = np.arctan2(0, home_x - x)        # heading toward home (y held at 0)
    new_x = x + speed * np.cos(angle)
    new_x = min(new_x, home_x)               # cap at destination
    heading = np.degrees(angle)
    return new_x, y, heading


# ── Find schoolmates ──────────────────────────────────────────────────────────

def find_schoolmates(agent_id, agents):
    focal = agents.loc[agents["id"] == agent_id].iloc[0]
    mates = agents[
        (agents["id"] != agent_id) &
        (agents["start_migration"]) &
        (agents["landward_migration"] == focal["landward_migration"])
    ]
    return mates


# ── Schooling (BOIDS rules) ───────────────────────────────────────────────────

def school_agent(agent_id, agents):
    focal = agents.loc[agents["id"] == agent_id].iloc[0]
    mates = find_schoolmates(agent_id, agents)

    if mates.empty:
        return focal["heading"], focal["speed"]

    # Nearest neighbor
    dists   = np.sqrt((mates["x"] - focal["x"])**2 +
                      (mates["y"] - focal["y"])**2)
    nn_dist = dists.min()
    nn_x    = mates.loc[dists.idxmin(), "x"]

    new_heading = focal["heading"]
    new_speed   = focal["speed"]

    if nn_dist < focal["minimum_separation"]:
        # Separate: slow down if neighbor is ahead, turn away otherwise
        if nn_x > focal["x"]:
            new_speed = max(focal["min_speed"], focal["speed"] - 0.1)
        else:
            new_heading = focal["heading"] + (
                focal["separate_coefficient"] / 100
            ) * np.sign(focal["x"] - nn_x) * 10

    else:
        # Cohere: turn toward centroid of schoolmates
        centroid_x  = mates["x"].mean()
        new_heading = focal["heading"] + (
            focal["cohere_coefficient"] / 100
        ) * np.sign(centroid_x - focal["x"]) * 10

        # Align: match mean heading of schoolmates
        mean_heading = mates["heading"].mean()
        new_heading  = new_heading + (
            focal["align_coefficient"] / 100
        ) * (mean_heading - new_heading)

    # Adjust speed: speed up if a schoolmate is faster, otherwise slow down
    if mates["speed"].max() > focal["speed"] + 0.1:
        new_speed = focal["speed"] * 1.1
    elif focal["speed"] > focal["min_speed"]:
        new_speed = focal["speed"] * 0.9
    new_speed = max(new_speed, focal["min_speed"])

    return new_heading % 360, new_speed
```

### Simulation loop

``` python
log_ticks = []

for tick in range(1, n_ticks + 1):

    # Step 1: migrate — face home and step forward
    migrating = (
        agents["start_migration"] &
        agents["landward_migration"] &
        ~agents["at_destination"]
    )

    for i in agents.index[migrating]:
        row = agents.loc[i]
        new_x, new_y, new_heading = migrate_landward(
            row["x"], row["y"], row["speed"], row["home_x"]
        )
        agents.at[i, "x"]       = new_x
        agents.at[i, "y"]       = new_y
        agents.at[i, "heading"] = new_heading

    # Step 2: school — separation, cohesion, alignment, speed adjustment
    school_results = [school_agent(aid, agents) for aid in agents["id"]]
    agents["heading"] = [r[0] for r in school_results]
    agents["speed"]   = [r[1] for r in school_results]

    # Step 3: check arrival
    agents["at_destination"]    = agents["x"] >= agents["home_x"]
    agents["landward_migration"] = ~agents["at_destination"]

    # Log summary statistics
    log_ticks.append({
        "tick"      : tick,
        "mean_x"    : agents["x"].mean(),
        "mean_speed": agents["speed"].mean(),
        "n_arrived" : agents["at_destination"].sum(),
    })

results = pd.DataFrame(log_ticks)
```

### Visualizing outputs

``` python
import matplotlib.pyplot as plt

fig, axes = plt.subplots(1, 3, figsize=(12, 4))

# Migration progress
axes[0].plot(results["tick"], results["mean_x"], color="#2c7bb6", linewidth=1.5)
axes[0].set_title("Mean agent position over time")
axes[0].set_xlabel("Tick")
axes[0].set_ylabel("Mean upstream position")

# Speed convergence from schooling
axes[1].plot(results["tick"], results["mean_speed"], color="#d7191c", linewidth=1.5)
axes[1].set_title("Mean agent speed over time")
axes[1].set_xlabel("Tick")
axes[1].set_ylabel("Mean speed")

# Cumulative arrivals
axes[2].step(results["tick"], results["n_arrived"],
             color="#1a9641", linewidth=1.5, where="post")
axes[2].set_title("Cumulative arrivals at home patch")
axes[2].set_xlabel("Tick")
axes[2].set_ylabel("Number of agents arrived")

plt.tight_layout()
plt.show()
```
